Why do languages tolerate heterography? An experimental investigation into the emergence of informative orthography

### 2024-01-15 manuscript/manuscript.tex
Add learnability analysis to ms

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
	\vspace*{2pt}
	\caption{Two models of heterography. \textbf{(A)} In the differentiation model, two meanings are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing. \textbf{(B)} In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model. Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
	\label{models_of_heterography}
	\end{figure}

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
	\vspace*{2pt}
	\caption{\textbf{(A)} Transmission-only procedure. Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items. These productions are then used as the training material for the next generation in the chain. The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent. \textbf{(B)} Transmission~+ Communication procedure. Each generation now consists of two participants who engage in a communicative task. Both participants receive training on the same system from the previous generation. Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
	\label{transmission}
	\end{figure}

@@

The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, following the same 3×3 layout used in Fig.~\ref{stimuli}. The color-coding of these matrices indicates similarity in suffix form: Similar colors are used to represent similar suffixes, making it easier to see how the suffixes pattern with meaning.\footnote{The color-coding was generated independently for each chain by selecting $n$ evenly-spaced hues, where $n$ is the number of unique forms that emerged across the chain, and mapping these hues onto the suffixes in alphabetical order.} For example, the system at Generation~9 in Chain~D uses three spellings (<koe>, <ko>, and <co>) to express the shape dimension, yielding a horizontal stripes pattern in the matrix representation. We describe such a system as ``redundant'' because shape was consistently and reliably expressed by the stem, so the suffix spelling system that emerged in this case conveys no additional information---the suffix simply repeats whichever shape was marked by the stem. Redundant suffix systems are characteristic of the Transmission-only condition, with similar outcomes occurring in Chains~C, E, F, and~H. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated towards transparency.

@@

@@

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_dif.eps}}
	\vspace*{2pt}
	\caption{\textbf{(A)} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange). The thin lines show each individual transmission chain (with a small amount of added jitter). The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data. Generation~0 is not included in the model, since it does not represent participant data. The black dotted lines show the communicative cost of the spoken language for comparison. In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure. \textbf{(B)} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
	\label{cost_dif}
	\end{figure}

@@

@@

Our first experiment tested whether informative, heterographic orthography could emerge through differentiation and whether it would emerge preferentially under communicative pressure. Overall, there was little evidence of systematic orthographic differentiation. Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <a> for pink, <b> for yellow, and <c> for blue). Such differentiation was, however, not forthcoming, even under communicative pressure. Instead, when spelling variation was conditioned on anything, it was usually conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure \parencite[e.g.,][]{Kirby:2015}.

So, what was different about the present experiment? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. Faced with variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.

It is particularly surprising that participants chose not to innovate novel systems in the communicative condition, despite there being a clear financial incentive to do so. Why might this be the case? Firstly, although innovation was not expressly forbidden in the instructions, it was perhaps suggested by the overall structure of the experiment that participants were expected to use the orthography learned during training. Participants who adopted this interpretation may have feared diverging from what they had learned, either because they themselves wanted to avoid breaking the rules or because they did not know whether their partner would feel comfortable breaking the rules. Secondly, even if participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners? They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems. These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space. Additionally, deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>). Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the intention? Differentiation is hard in the written modality.

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_con.eps}}
	\vspace*{2pt}
	\caption{\textbf{(A)} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange). The thin lines show each individual transmission chain (with a small amount of added jitter). The thick lines and shaded regions show the mean and 90\% HDI of a regression model fit to the data. Generation~0 is not included in the model, since it does not represent participant data. The black dotted lines show the communicative cost of the spoken language, which increases with epoch. In general, the orthographic systems tend to remain more informative under communicative pressure. \textbf{(B)} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
	\label{cost_con}
	\end{figure}

@@

>>>>

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/models_of_heterography.eps}}
	\vspace*{2pt}
	\caption{Two models of heterography. \textbf{A} In the differentiation model, two meanings are, at time $T_1$, expressed by a single phonetic form $P_1$ and a single orthographic form $O_1$; however, by time $T_2$, two orthographic forms have emerged to differentiate the meanings in writing. \textbf{B} In the conservation model, the two distinct phonetic forms that existed at time $T_1$ have become homophonous by time $T_2$, but the two corresponding orthographic forms have been conserved, resulting in the same state of heterography as in the differentiation model. Adapted from \textcite[pp.~325--326]{Berg:2021} with permission.}
	\label{models_of_heterography}
	\end{figure}

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/transmission.eps}}
	\vspace*{2pt}
	\caption{\textbf{A} Transmission-only procedure. Each generation consists of a single participant, who first receives training on six of the nine suffixes and then produces suffixes for all nine items. These productions are then used as the training material for the next generation in the chain. The initial system of suffixes is randomly generated with high spelling variation; but by the ninth generation, the system is expected to become transparent. \textbf{B} Transmission~+ Communication procedure. Each generation now consists of two participants who engage in a communicative task. Both participants receive training on the same system from the previous generation. Under a communicative pressure, the suffix spellings are expected to become expressive of color, despite the spoken forms being homophonous.}
	\label{transmission}
	\end{figure}

@@

The results from all ten chains (labeled A--J) in the Transmission-only condition are shown in Fig.~\ref{dif_lrn}. Each 3×3 matrix represents the suffix spelling system in use at a particular generation with shape represented along the rows and color represented along the columns, following the same 3×3 layout used in Fig.~\ref{stimuli}. The color-coding of these matrices indicates similarity in suffix form: Similar colors are used to represent similar suffixes, making it easier to see how the suffixes pattern with meaning.\footnote{The color-coding was generated independently for each chain by selecting $n$ evenly-spaced hues, where $n$ is the number of unique forms that emerged across the chain, and mapping these hues onto the suffixes in alphabetical order.} For example, the system at Generation~9 in Chain~D uses three spellings (<koe>, <ko>, and <co>) to express the shape dimension, yielding a horizontal stripes pattern in the matrix representation. We describe such a system as ``redundant'' because shape was consistently and reliably expressed by the stem, so the suffix spelling system that emerged in this case conveys no additional information---the suffix simply repeats whichever shape was marked by the stem. Redundant suffix systems are characteristic of the Transmission-only condition, with similar outcomes occurring in Chains~C, E, F, and~H. We also saw the emergence of fully transparent suffix spelling systems in Chains~A, B, I, and~J. Chain~I, for example, ultimately uses a single spelling, <coe>, to represent the /kəʊ/ sound; that is, the written suffix forms make no distinction between shapes or colors, just like the spoken language. Chain~G did not settle on a clear pattern, using <co>, <coe>, and <coh> somewhat interchangeably in the final generation, although there were some signs of a color-expressive system emerging in, for example, Generation~6, where yellow is consistently spelled <coh>, blue is consistently spelled <coe>, and pink uses both <co> and <coe>. The only other signs of color-expressive systems are H1, which was rapidly converted into a redundant system in subsequent generations, and J2, which ultimately degenerated toward transparency.

@@

\subsubsection{Typological distribution}

@@

\subsubsection{Communicative cost}

@@

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_dif.eps}}
	\vspace*{2pt}
	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange). The thin lines show each individual transmission chain (with a small amount of added jitter). The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data. Generation~0 is not included in the model, since it does not represent participant data. The black dotted lines show the communicative cost of the spoken language for comparison. In general, the orthographic systems tend to become more costly over time, but less so under communicative pressure. \textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
	\label{cost_dif}
	\end{figure}

@@

\subsubsection{Learnability}

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/te_dif.eps}}
	\vspace*{2pt}
	\caption{\textbf{A} Transmission error by generation under Transmission-only (dark green) and Transmission~+ Communication (orange). The thin lines show each individual transmission chain (with a small amount of added jitter). The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data. Generation~0 is not included in the model, since it does not represent participant data. Transmission error tends to decrease over time in both conditions, suggesting that the orthographic systems become more learnable; however, transmission in Transmission~+ Communication tends to remain a little higher. \textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
	\label{te_dif}
	\end{figure}

Aside from the informativeness of the orthographic systems, we also predicted that the systems would become easier to learn over time in both conditions, albeit for slightly different reasons. In Transmission-only, the orthographic system is expected to become increasingly learnable as it degenerates into a single, transparent suffix form. In Transmission~+ Communication, the system is expected to become more learnable as the unsystematic, holistic systems transform into other easier to learn systems (notably expressive systems, although---as noted already---these were not particularly forthcoming). We operationalize learnability as transmission error---the amount of error that the participant at Generation~$i$ made in reproducing the orthographic system that existed at Generation~$i-1$. Following other papers, we define transmission error as the mean Levenshtein edit distance between the corresponding orthographic forms in consecutive generations.

The results are plotted in Fig.~\ref{te_dif}, along with the posterior parameter estimates of a regression model fit to the data (this is the same model outlined above, except that we use a 2nd degree polynomial which better characterized the data). In Transmission-only, transmission error tends toward zero as the systems become trivial to learn and reproduce. In Transmission~+ Communication, there is some slight evidence for an uptick in transmission error in the last few generations as participants start to introduce novel disambiguating forms (notably English color letters), which diverge from the previous generation. Although there is little evidence to suggest that the shapes of the curves differ between conditions, there is a relatively clear difference in the intercepts: Transmission error remains higher in the communicative condition. This is to be expected because the goal in the Transmission-only condition is to reproduce the forms taught in training, whereas the goal in the communicative condition is to devise a system that permits accurate communication, which necessitates some change to the system. Nevertheless, the effect size is rather small (STATS), suggesting again that the presence of the communicative task did not have a big impact on how the systems evolved.

@@

Our first experiment tested whether informative, heterographic orthography could emerge through differentiation and whether it would emerge preferentially under communicative pressure. Overall, there was little evidence of systematic orthographic differentiation. Other than resorting to English, such differentiation could have been achieved in a number of ways, most obviously through the conditioning of spelling variation on meaning (e.g., <k> for pink, <c> for yellow, and <q> for blue), but also through less obvious strategies such as the use of length (e.g., <o> for pink, <oo> for yellow, and <ooo> for blue) or the use of arbitrary silent letters (e.g., <x> for pink, <f> for yellow, and <m> for blue). Such differentiation was not forthcoming, however, even under communicative pressure. Instead, if spelling variation was conditioned on anything, it was conditioned on shape, resulting in redundant suffix spellings. This result is in stark contrast to most prior experimental iterated learning studies, in which informative, compositional systems do typically emerge, especially under communicative pressure \parencite[e.g.,][]{Kirby:2015}.

So, what was different about the present experiment compared the large body of prior experimental iterated learning studies? The primary difference was the presence of a spoken language that is decidedly \textit{not} informative about one of the dimensions. Indeed, the point of our experiment is to see whether orthography can resist phonology under sufficient pressure for informativeness. The presence of homophonous suffix forms acts as a cue to participants that the language itself does not mark color and that, therefore, the orthography should also not mark color. Faced with orthographic variation that could be conditioned on either dimension, learners appear to rule out the possibility that it might be conditioned on color, since such a hypothesis would be in conflict with the spoken language. As a result, any variation in spelling comes to be associated with particular stems, resulting in the emergence of redundant suffix spellings that serve no real purpose. A rough analog of this redundancy can be found in English in the spelling of /-ʃən/ (<cian>, <cion>, <sion>, <ssion>, or <tion>), which is conditioned on the stem (e.g., \textit{magician} \textit{suspicion}, \textit{expulsion}, \textit{transmission}, and \textit{station}) following a complex set of rules \parencite[pp.~420--421]{Carney:1994}.

It is particularly surprising that participants chose not to innovate novel systems in the communicative condition, despite there being a clear financial incentive to do so. Why might this be the case? Firstly, although innovation was not expressly forbidden in the instructions, it is possible that some participants may have inferred such a rule. Participants who adopted such an interpretation may have feared diverging from what they had learned, either because they themselves wanted to avoid breaking these perceived rules or because they did not know whether their partner would feel comfortable breaking them. Secondly, even when participants were willing to diverge, they were presented with an alignment problem: How should they decide what system of color marking to use when there is no channel through which they can agree a system with their partners? They can only use trial-and-error, and even then, the process is complicated by participants simultaneously adapting to each other's systems. These are dynamics that also exist in the real world, especially in writing where the reader and writer are separated in time and space. Additionally, deviation from spelling norms is often considered socially unacceptable or uneducated, even when novel spellings are advantageous (e.g., the non-standard spelling <thru> has struggled for decades to make any headway against <through>). Even under more progressive attitudes to change and innovation, the problem of achieving community buy-in remains, especially when the change is not imposed by top-down authority: If someone decided to start using the spelling <banque> to differentiate the financial institution from river banks, would anyone take that spelling seriously or even understand the intention? Differentiation is hard in the written modality.

@@

\subsubsection{Communicative cost}

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/cost_con.eps}}
	\vspace*{2pt}
	\caption{\textbf{A} Communicative cost by generation under Transmission-only (dark green) and Transmission~+ Communication (orange). The thin lines show each individual transmission chain (with a small amount of added jitter). The thick lines and shaded regions show the mean and 90\% HDI of a regression model fit to the data. Generation~0 is not included in the model, since it does not represent participant data. The black dotted lines show the communicative cost of the spoken language, which increases with epoch. In general, the orthographic systems tend to remain more informative under communicative pressure. \textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
	\label{cost_con}
	\end{figure}

@@

\subsubsection{Learnability}

	\begin{figure}
	\makebox[\textwidth][c]{\includegraphics[]{figs/te_con.eps}}
	\vspace*{2pt}
	\caption{\textbf{A} Transmission error by generation under Transmission-only (dark green) and Transmission~+ Communication (orange). The thin lines show each individual transmission chain (with a small amount of added jitter). The thick lines and shaded regions show the mean and 90\% HDI of a polynomial regression model fit to the data. Generation~0 is not included in the model, since it does not represent participant data. Transmission error tends to decrease over time in both conditions, suggesting that the orthographic systems become more learnable; however, transmission in Transmission~+ Communication tends to remain a little higher. \textbf{B} Posterior parameter estimates from the regression model, along with the posterior differences across conditions.}
	\label{te_con}
	\end{figure}

Fig.~\ref{te_con} plots results for transmission error following the same methods outlined previously. Overall there were no strong differences between conditions, although we did not particularly expect to find any. Descriptively, in Transmission-only transmission error follows an inverted-\textsc{u} shape; it first increases in response to the homophony introduced at Generation~4, but then decreases as the orthographic systems converge on easy to learn patterns. In Transmission~+ Communication, the transmission error tend to remain fairly stable through Epochs~II and~III, representing a kind of baseline level of error in response to homophony.
